Bayes Optimal Informer Sets for Drug Discovery\thanks{abc}

### 2020-04-15 BOISE/Paper_draft/draft.tex
draft update

\begin{titlepage}
\title{Bayes Optimal Informer Sets for Drug Discovery\thanks{abc}}
\author{Peng Yu, {\sc contributing members of UW drug discovery group}, \\  \& Michael A. Newton\thanks{abc}}
\date{\today}
\maketitle
\begin{abstract}
In early stage drug discovery, there has been significant recent effort in developing compound prioritization methods for a new target from minimal data. One kind of such methods is informer based ranking (IBR) methods, which select an initial subset of compounds, so-called informer set, to be screened based on original data, and rank additional compounds for the new target based on this additional screening data and original data. However, the selection of informer set in previous methods is often heuristic, based on chemical diversity or randomized algorithms with no statistical guarantees. In this article, we  formalize the informer set ranking procedure as a two-stage decision problem and introduce a new method named as Bayes Optimal Informer SEt (BOISE) to select informer set as well as ranking the compounds against the new target. BOISE is a fully Bayesian procedure that selects informer set based on posterior expected loss and ranks compounds based on posterior expectation. It is guaranteed to be Bayes rule under proposed model and loss if computed exactly, and we propose a computationally efficient version of BOISE that achieves a balance between ranking performance and computation complexity. Both proposed methods and previous methods are applied on human kinase chemogenomics data set and anti-cancer drug sensitivity data set and evaluated under various ranking criteria. We demonstrate that proposed methods are significantly better than previous methods under all metrics. 
\vspace{0in}\\
\noindent\textbf{Keywords:} key1, key2, key3\\
\vspace{0in}\\
\noindent\textbf{JEL Codes:} key1, key2, key3\\

@@

One common misunderstanding in informer based ranking method is that someone will tend to select the most frequent hitters in bioactivity matrix as the informer set. We come up with a simple counterexample in Appendix to show that informer set ranking problem is a more complicated problem. In short, most informative compounds are not those that are most likely to be active. Zhang, Erichsen and Lee {\em et al.} (2019) introduced several informer based ranking strategies and compared
them empirically in data on protein kinases. Though differing in detail,
the proposed strategies all partition the initial
bioactivity data, producing  clusters of relatively similar targets; 
then informer compounds are  those that are predictive of the cluster label of a target.  Among kinase targets, the proposed bioactivity-based strategies had better predictive
performance than commonly used strategies based upon chemical
structure.   Since effective informer based ranking strategies leverage statistical
patterns in the initial bioactivity data, we reason that
statistical modeling may provide a useful approach to deriving
more effective IBR strategies than are currently available.

@@

Now let $i^* \notin I$ index a new protein target of interest, on which we have no initial bioactivity data.   
We are asked to find a so-called informer set $A \subset J$ using initial data $x_0$. The size of $A$ is a
given small number $n_A \ll n$. We will perform
a new experiment on new target $i^*$ and informer set $A$ to obtain intermediate data 
\begin{eqnarray*}
x_{A} = \left\{ x_{i^*j}: j \in A \right\}.
\end{eqnarray*}
This $x_A$ measure the bioactivity of informer
set compounds against the new target $i^*$.  The initial and intermediate
data are then used to prioritize compounds for further testing.
To formalize this, let us suppose that we must select a final
top set $T$, of certain cardinality $n_T$, with
\begin{eqnarray*}
T = T\left( x_0, A, x_A \right) \subset J.
\end{eqnarray*}
The notation is intended to emphasize that the proposed top set $T$ is constructed
from initial data, the informer set, and the intermediate data.
We imagine that all the compounds $j$ in $T$ will be
tested against new target $i^*$, and we seek this  {\em top set} $T$ to have as many
active compounds as possible.

@@

Our statistical model is a flexible generative model on both original known data $x_0$ and new unknown data $x_{i^*}$. 
To represent similarity among targets, we assume there is a clustering structure $\mathcal{C}$ on targets $i\in I$, and targets within each cluster will have similar interactions with compounds.

To be more specific, we assume $\mathcal{C} = \{c_k, k = 1,2,\cdots\}$ is a partition on all $m$ targets. Each $c_k$ labels $m_k$ targets out of total $m$ targets. New target $i^*$ may be in one existing cluster $c_k$ or form a new cluster by itself, denoted as $c_0$. For each entry $x_{ij}$ of the bioactivity data $x_0$, we assume that it is a realization of mutually independent Bernoulli trails,  determined by both compound $j$ and the cluster $c_k$ where corresponding target $i$ belongs:

@@

\begin{figure}[!ht]
\caption{\label{fig:pkis8} 
{\bf Violin plot comparison of different methods under informer size $n_A =8$ with respect to compound ranking performance assessed by NEF10 and ROCAUC.} Each method is evaluated through leave one target out cross validation of PKIS1 targets. NEF10 is a measurement of active compounds enrichment in top $10\%$ ranked compounds, after normalization with regard to maximum theoretical enrichment that could be achieved at the 10\% threshold for the target of interest. NEF10 and ROCAUC of 0.5 indicates a random guess, and of 1.0 represents ideal ranking with all active compounds prioritized above the inactives. The median and interquartile ranges are displayed as a white circle and black bars, respectively.}
\centering
\includegraphics[width=5.0in]{PKIS1_8_eval.png}
\end{figure}

Same thing happens when we boost informer set size from $n_A = 8$ to $16$. Since BOISE takes too much time for a leave-one-out cross validation under $n_A=16$, we can not afford it, instead we treat Accelerated-BOISE as a lower bound of BOISE performance as Figure \ref{fig:pkis8} tell us. Empirical results summarized in Figure \ref{fig:pkis16} show that AC-BOISE is still significantly better than previous methods: Under NEF10 metric, Adaptive Selection (AS16) is the best method other than AC-BOISE and the paired t-test shows that their mean difference is $0.0415$ with p-value $=6.199\times 10^{-7}$;  Under ROCAUC metric, Regression Selection (RS16) is the best method other than AC-BOISE and the paired t-test shows that their mean difference is $0.0294$ with p-value $=1.303\times 10^{-4}$. 
\begin{figure}[!ht]
\caption{\label{fig:pkis16} 
{\bf Violin plot comparison of different methods under informer size $n_A =16$ with respect to compound ranking performance assessed by NEF10 and ROCAUC.}} 
\centering
\includegraphics[width=5.0in]{PKIS1_16_eval.png}
\end{figure}

@@

\begin{figure}[!ht]
\caption{\label{fig:mcc} 
{\bf Violin plot comparison of different methods under informer size $n_A =8$ with respect to performance assessed by Matthews Correlation Coefficient (MCC) and F1 Score.} MCC takes value in $[-1,1]$ and F1 Score takes value in $[0,1]$. MCC of $0$ and F1 Score of $0.5$ indicate a random guess; MCC and F1 Score of 1.0 represent ideal ranking with all active compounds prioritized above the inactives. The median and interquartile ranges are displayed as a white circle and black bars, respectively.} 
\centering
\includegraphics[width=5.0in]{PKIS1_8_mccf1.png}
\end{figure}

@@

\begin{figure}[!ht]
\caption{\label{fig:gdsc} 
{\bf Violin plot comparison of different methods under informer sizes $n_A =8$ and $n_A = 16$ on GDSC data set with respect to compound ranking performance assessed by NEF10 and ROCAUC.} Each method is evaluated through leave one out cross validation of GDSC cancer cell lines. The median and interquartile ranges are displayed as a white circle and black bars, respectively.}
\centering
\includegraphics[width=5.0in]{GDSC.png}
\end{figure}

@@

\begin{figure}[!ht]
\caption{\label{fig:mcmc} 
{\bf Mixing properties of MCMC sampling over clusters on PKIS1 and GDSC} The plots are based on the chain over numbers of clusters in each DPMM sample.}
\centering
\includegraphics[width=5.0in]{DPMM_performance.png}
\end{figure}

>>>>

\begin{titlepage}
\title{Bayes Optimal Informer Sets for Drug Discovery\thanks{abc}}
\author{Peng Yu, {\sc contributing members of UW drug discovery group}, \\  \& Michael A. Newton\thanks{abc}}
\date{\today}
\maketitle
\begin{abstract}
In early stage drug discovery, there has been significant recent effort in developing compound prioritization methods for a new target from minimal data. One kind of such methods is informer based ranking (IBR) methods, which selects an initial subset of compounds, so-called informer set, to be screened based on original data, and ranks additional compounds for the new target based on this additional screening data and original data. However, the selection of informer set in previous methods is often heuristic, based on chemical diversity or randomized algorithms with no statistical guarantees. In this article, we  formalize the informer set ranking procedure as a two-stage decision problem and introduce a new method named as Bayes Optimal Informer SEt (BOISE) to select informer set as well as rank the compounds against the new target. BOISE is a fully Bayesian procedure that selects informer set based on posterior expected loss and ranks compounds based on posterior expectation. It is guaranteed to be Bayes rule under proposed model and loss if computed exactly, and we propose a computationally efficient version of BOISE that achieves balance between ranking performance and computation complexity. Both proposed methods and previous methods are applied on human kinase chemogenomics data set and anti-cancer drug sensitivity data set and evaluated under various ranking criteria. We demonstrate that proposed methods are significantly better than previous methods under all metrics. 
\vspace{0in}\\
\noindent\textbf{Keywords:} key1, key2, key3\\
\vspace{0in}\\
\noindent\textbf{JEL Codes:} key1, key2, key3\\

@@

One common misunderstanding of informer based ranking method is that many researchers tend to select the most frequent hitters in bioactivity matrix as the informer set. We come up with a simple counterexample in Appendix to show that informer set ranking problem is a more complicated problem. In short, most informative compounds are not those that are most likely to be active. Zhang, Ericksen and Lee {\em et al.} (2019) introduced several informer based ranking strategies and compared
them empirically in data on protein kinases. Though differing in detail,
the proposed strategies all partition the initial
bioactivity data, producing  clusters of relatively similar targets; 
then informer compounds are  those that are predictive of the cluster label of a target.  Among kinase targets, the proposed bioactivity-based strategies had better predictive
performance than commonly used strategies based upon chemical
structure.   Since effective informer based ranking strategies leverage statistical
patterns in the initial bioactivity data, we reason that
statistical modeling may provide a useful approach to deriving
more effective IBR strategies than are currently available.

@@

Now let $i^* \notin I$ index a new protein target of interest, on which we have no initial bioactivity data.   
We are asked to find a so-called informer set $A \subset J$ using initial data $x_0$. The size of $A$ is a
given small number $n_A \ll n$. We will perform
a new experiment on new target $i^*$ and informer set $A$ to obtain intermediate data 
\begin{eqnarray*}
x_{A} = \left\{ x_{i^*j}: j \in A \right\}.
\end{eqnarray*}
This $x_A$ measure the bioactivity of informer
set compounds against the new target $i^*$.  The initial and intermediate
data are then used to prioritize compounds for further testing.
To formalize this, let us suppose that we must select a final
top set $T$, of certain cardinality $n_T$, with
\begin{eqnarray*}
T = T\left( x_0, A, x_A \right) \subset J.
\end{eqnarray*}
The notation is intended to emphasize that the proposed top set $T$ is constructed
from initial data, the informer set, and the intermediate data.
We imagine that all the compounds $j$ in $T$ will be
tested against new target $i^*$, and we seek this  top set $T$ to have as many
active compounds as possible.

@@

Our statistical model is a Bayesian model on both original known data $x_0$ and new target data $x_{i^*}$. 
To represent similarity among targets, we assume there is a clustering structure $\mathcal{C}$ on targets $i\in I$, and targets within each cluster will have similar interactions with compounds. 
To be more specific, we assume $\mathcal{C} = \{c_k, k = 1,2,\cdots\}$ is a partition on all $m$ targets. Each $c_k$ labels $m_k$ targets out of total $m$ targets. New target $i^*$ may be in one existing cluster $c_k$ or form a new cluster by itself, denoted as $c_0$. For each entry $x_{ij}$ of the bioactivity data $x_0$, we assume that it is a realization of mutually independent Bernoulli trails,  determined by both compound $j$ and the cluster $c_k$ where corresponding target $i$ belongs:

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=5.0in]{PKIS1_8_eval.png}
\caption{\label{fig:pkis8} 
{\bf Violin plot comparison of different methods under informer size $n_A =8$ with respect to compound ranking performance assessed by NEF10 and ROCAUC.} Each method is evaluated through leave one target out cross validation of PKIS1 targets. NEF10 is a measurement of active compounds enrichment in top $10\%$ ranked compounds, after normalization with regard to maximum theoretical enrichment that could be achieved at the 10\% threshold for the target of interest. NEF10 and ROCAUC of 0.5 indicates a random guess, and of 1.0 represents ideal ranking with all active compounds prioritized above the inactives. The median and interquartile ranges are displayed as a white circle and black bars, respectively.}
\end{figure}

Same thing happens when we boost informer set size from $n_A = 8$ to $16$. Since BOISE takes too much time for a leave-one-out cross validation under $n_A=16$, we can not afford it, instead we treat Accelerated-BOISE as a lower bound of BOISE performance as Figure \ref{fig:pkis8} tell us. Empirical results summarized in Figure \ref{fig:pkis16} show that AC-BOISE is still significantly better than previous methods: Under NEF10 metric, Adaptive Selection (AS16) is the best method other than AC-BOISE and the paired t-test shows that their mean difference is $0.0415$ with p-value $=6.199\times 10^{-7}$;  Under ROCAUC metric, Regression Selection (RS16) is the best method other than AC-BOISE and the paired t-test shows that their mean difference is $0.0294$ with p-value $=1.303\times 10^{-4}$. 
\begin{figure}[!ht]
\centering
\includegraphics[width=5.0in]{PKIS1_16_eval.png}
\caption{\label{fig:pkis16} 
{\bf Violin plot comparison of different methods under informer size $n_A =16$ with respect to compound ranking performance assessed by NEF10 and ROCAUC.}}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=5.0in]{PKIS1_8_mccf1.png}
\caption{\label{fig:mcc} 
{\bf Violin plot comparison of different methods under informer size $n_A =8$ with respect to performance assessed by Matthews Correlation Coefficient (MCC) and F1 Score.} }
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=5.0in]{GDSC.png}
\caption{\label{fig:gdsc} 
{\bf Violin plot comparison of different methods under informer sizes $n_A =8$ and $n_A = 16$ on GDSC data set with respect to compound ranking performance assessed by NEF10 and ROCAUC.} Each method is evaluated through leave one out cross validation of GDSC cancer cell lines. The median and interquartile ranges are displayed as a white circle and black bars, respectively.}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=5.0in]{DPMM_performance.png}
\caption{\label{fig:mcmc} 
{\bf Mixing properties of MCMC sampling over clusters on PKIS1 and GDSC} The plots are based on the chain over numbers of clusters in each DPMM sample.}
\end{figure}
